---
title: "Real-World Machine Learning" 
subtitle: "Chapter 3"
author: "Paul Adamson"
date: "December 7, 2016"
output: html_document
---

This file contains R code to accompany Chapter 3 of the book 
["Real-World Machine Learning"](https://www.manning.com/books/real-world-machine-learning),
by Henrik Brink, Joseph W. Richards, and Mark Fetherolf.  The code was contributed by
[Paul Adamson](http://github.com/padamson). 

*NOTE: working directory should be set to this file's location.*

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(dplyr)
library(vcd)
library(AppliedPredictiveModeling)
library(caret)
library(ellipse)
library(kknn)
library(gridExtra)
library(grid)
library(randomForest)
set.seed(3456)
.pardefault <- par()
```

## Figure 3.4 A subset of the Titanic Passengers dataset

We are going to be interested in predicting survival, so it is useful to specify 
the `Survived` variable to be of type `factor`. For visualizing the data, 
it is also useful to use the `revalue` function to specify the `no` and `yes`
levels for the `factor` variable. The `kable` function is built into the `knitr`
package.

```{r figure3.4, cache=TRUE}
titanic <- read.csv("../data/titanic.csv", 
                    colClasses = c(
                      Survived = "factor",
                      Name = "character",
                      Ticket = "character",
                      Cabin = "character"))
titanic$Survived <- revalue(titanic$Survived, c("0"="no", "1"="yes"))
kable(head(titanic, 6), digits=2)
```

## Figure 3.5 Mosaic plot for Titanic data: Gender vs. survival

The ["Visualizing Categorical Data" package  (`vcd`)](https://cran.r-project.org/web/packages/vcd/vcd.pdf)
provides an excellent set of functions for exploring categorical data,
including mosaic plots.

```{r figure3_5, cache=TRUE, dependson="figure3.4"}
mosaic(
  ~ Sex + Survived,
  data = titanic, 
  main = "Mosaic plot for Titanic data: Gender vs. survival",
  shade = TRUE,
  split_vertical = TRUE,
  labeling_args = list(
    set_varnames = c(
      Survived = "Survived?",
      Sex = "Gender")))
```

## Figure 3.6 Processed data

First, we get rid of the variables that we do not want in our model.
(`Cabin` might actually be useful, but it's not used here.)
Then we use `is.na` to set missing age values to -1.
The `mutate` and `select` functions make it easy to take square root of 
the `Fare` variable and then drop it from the dataset.
We then drop rows with missing `Embarked` data and remove the unused level 
`""`. 
Finally, we convert `factor` variables to dummy variables using the 
`dummyVars` function in the `caret` package.
To avoid perfect collinearity (a.k.a. the dummy variable trap), we set
the `fullRank` parameter to `TRUE`.  `Survived.yes` is then converted back
to a `factor` variable.

```{r figure3.6, cache=TRUE, dependson="figure3.4"}
titanicTidy <- subset(titanic, select = -c(PassengerId, Name, Ticket, Cabin))

titanicTidy$Age[is.na(titanicTidy$Age)] <- -1

titanicTidy <- titanicTidy %>%
  mutate(sqrtFare = sqrt(Fare)) %>%
  select(-Fare)

titanicTidy <- titanicTidy %>%
  filter(!(Embarked=="")) %>%
  droplevels

dummies <- dummyVars(" ~ .", data = titanicTidy, fullRank = TRUE)
titanicTidyNumeric <- data.frame(predict(dummies, newdata = titanicTidy))

titanicTidyNumeric$Survived.yes <- factor(titanicTidyNumeric$Survived.yes)
kable(head(titanicTidyNumeric))
```

## Figure 3.10 Four randomly chosen handwritten digits from the MNIST database

Thanks to [Longhow Lam](https://longhowlam.wordpress.com/2015/11/25/a-little-h2o-deeplearning-experiment-on-the-mnist-data-set/)
for posting the code used in the `displayMnistSamples` function that display's 
digits from the MNIST dataset.

```{r figure3.10, cache=TRUE,fig.height=2}
mnist <- read.csv("../data/mnist_small.csv",
                  colClasses = c(label = "factor"))
displayMnistSamples <- function(x) {
  for(i in x){
  y = as.matrix(mnist[i, 2:785])
  dim(y) = c(28, 28)
  image( y[,nrow(y):1], axes = FALSE, col = gray(0:255 / 255))
  text( 0.2, 0, mnist[i,1], cex = 3, col = 2, pos = c(3,4))
  }
}
par( mfrow = c(1,4), mai = c(0,0,0,0.1))
displayMnistSamples(sample(1:length(mnist),4))
```

## Figure 3.11 Table of predicted probabilities from a k-nearest neighbors classifier, as applied to the MNIST dataset

```{r figure3.11, cache=TRUE, dependson="figure3.10"}
trainIndex <- createDataPartition(mnist$label, p = .8, 
                                  list = FALSE, 
                                  times = 1)
mnistTrain <- mnist[ trainIndex,]
mnistTest  <- mnist[-trainIndex,]

mnist.kknn <- kknn(label~., mnistTrain, mnistTest, distance = 1,
                   kernel = "triangular")

confusionMatrix(fitted(mnist.kknn),mnistTest$label)

rowsInTable <- 1:10
prob <- as.data.frame(mnist.kknn$prob[rowsInTable,])
mnistResultsDF <- data.frame(mnistTest$label[rowsInTable],
                             mnist.kknn$fit[rowsInTable],
                             prob)


kable(mnistResultsDF, digits=2,
      col.names=c("actual","fit",0:9))
```

## Figure 3.13 Small subset of the Auto MPG data

```{r figure3.13, cache=TRUE}
auto <- read.csv("../data/auto-mpg.csv",
                 colClasses = c(
                      origin = "factor"))

auto$origin <- revalue(auto$origin, 
                       c("1\t"="USA", "2\t"="Europe", "3\t"="Asia"))

kable(head(auto,5))
```

## Figure 3.14 Scatter plots of Vehicle Weight and Model Year versus MPG

```{r figure3.14, cache=TRUE, dependson="figure3.13", warning=FALSE}
par(.pardefault)
p1<-ggplot(auto, aes(weight, mpg)) + 
  geom_point() +
  labs(y = "Miles per gallon",
       x = "Vehicle weight")
p2<-ggplot(auto, aes(modelyear, mpg)) + 
  geom_point() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  labs(x = "Model year")
grid.arrange(p1,p2,ncol=2, 
             top=textGrob("Scatterplots for MPG data",
                          gp=gpar(fontsize=14,font=8)))
```

## Figure 3.15 The Auto MPG data after expanding the categorical Origin column

Note that the row numbering differs between python and R by 1
(python starts row numbering at 0, and R starts at 1).

```{r figure3.15, cache=TRUE, dependson="figure3.13"}
dummies <- dummyVars(" ~ .", data = auto, fullRank = TRUE)
autoNumeric <- data.frame(predict(dummies, newdata = auto))

kable(tail(autoNumeric,5))
```

## Figure 3.16 Comparing MPG predictions on a held-out testing set to actual values

```{r figure3.16, cache=TRUE, dependson="figure3.15"}
trainIndex <- createDataPartition(autoNumeric$mpg, p = .8, 
                                  list = FALSE, 
                                  times = 1)

autoTrain <- autoNumeric[ trainIndex,]
autoTest  <- autoNumeric[-trainIndex,]

lmFit <- train(mpg ~ ., data = autoTrain,
               method = "lm")
lmPred <- predict(lmFit, newdata = autoTest)

rowsInTable <- 1:5
kable(data.frame("Origin.Europe" = autoTest$origin.Europe[rowsInTable],
                 "Origin.Asia" = autoTest$origin.Asia[rowsInTable],
                 "MPG" = autoTest$mpg[rowsInTable],
                 "Predicted MPG" = lmPred[rowsInTable]))
```

## Figure 3.17 A scatter plot of the actual versus predicted values on the held-out test set. The diagonal line shows the perfect regressor. The closer all of the predictions are to this line, the better the model.

```{r figure3.17, cache=TRUE, dependson="figure3.16"}
ggplot(autoTest, aes(x=mpg, y=lmPred)) + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0) +
  labs(x="MPG", y="Predicted MPG")
```

## Figure 3.18 Table of actual versus predicted MPG values for the nonlinear random forest regression model

```{r figure3.18, cache=TRUE, dependson="figure3.16"}
rfFit <- train(mpg ~ ., data = autoTrain,
               method = "rf")
rfPred <- predict(rfFit, newdata = autoTest)

kable(data.frame("Origin.Europe" = autoTest$origin.Europe[rowsInTable],
                 "Origin.Asia" = autoTest$origin.Asia[rowsInTable],
                 "MPG" = autoTest$mpg[rowsInTable],
                 "Predicted MPG" = rfPred[rowsInTable]))

```

## Figure 3.19 Comparison of MPG data versus predicted values for the nonlinear random forest regression model

```{r figure3.19, cache=TRUE, dependson="figure3.18"}
ggplot(autoTest, aes(x=mpg, y=rfPred)) + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0) +
  labs(x="MPG", y="Predicted MPG")
```
