---
title: "Real-World Machine Learning"
subtitle: "Chapter 2"
author: "Paul Adamson"
date: "December 7, 2016"
output: html_document
---

This file contains R code to accompany Chapter 2 of the book 
["Real-World Machine Learning"](https://www.manning.com/books/real-world-machine-learning),
by Henrik Brink, Joseph W. Richards, and Mark Fetherolf.  The code was contributed by
[Paul Adamson](http://github.com/padamson). 

*NOTE: working directory should be set to this file's location.*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vcd)
library(sm)
```

## Listing 2.1 Convert categorical features to numerical binary features

Creating dummy variables in R is extremely easy with the `model.matrix` function.
In the below code, when the `personData` dataframe is created,
the `maritalstatus` variable is of type `factor` with levels "single" and
"married" by default. 

```{r listing2.1}
personData <- data.frame(person = 1:2, 
                         name = c("Jane Doe", "John Smith"),
                         age = c(24, 41),
                         income = c(81200, 121000),
                         maritalstatus = c("single","married"))
personDataNew <- data.frame(personData[,1:4], 
                            model.matrix(~ maritalstatus - 1, 
                                         data = personData)) 
str(personData)
```

In the call to `model.matrix`, the −1 in the model formula 
ensures that we create a dummy variable for each of the two marital statuses 
(technically, it suppresses the creation of an intercept). 
```{r}
model.matrix(~ maritalstatus - 1, 
             data = personData)
```
The matrix of dummy variables is then joined to the original dataframe (minus
the maritalstatus column) with another call to `data.frame`.

## Listing 2.2 Simple feature extraction on Titanic cabins

The packages [`dplyr`](https://cran.rstudio.com/web/packages/dplyr) and 
[`tidyr`](https://cran.r-project.org/web/packages/tidyr) are excellent 
for tidying and preprocessing data, 
including creating new features from existing ones. (Note: `plyr` will
be used later, but we must load it prior to `dplyr`.)

```{r listing2.2}
titanic <- read.csv("../data/titanic.csv", 
                    colClasses = c(
                      Survived = "factor",
                      Name = "character",
                      Ticket = "character",
                      Cabin = "character"))

titanic$Survived <- revalue(titanic$Survived, c("0"="no", "1"="yes"))

titanicNew <- titanic %>%
  separate(Cabin, into = "firstCabin", sep = " ", extra = "drop", remove = FALSE) %>%
  separate(firstCabin, into = c("cabinChar", "cabinNum"), sep = 1) %>%
  rowwise() %>%
  mutate(numCabins = length(unlist(strsplit(Cabin, " "))))

str(titanicNew)
```

In Listing 2.2, `read.csv` is used to read in the comma separated value
(csv) data file. The `colClasses` argument is used to specify the correct
class for some features. Then, the `revalue` function changes the levels
of the `Survived` factor variable so that '0' indicates 'no' and '1'
indicates 'yes'. The `titanicNew` dataframe is then created by piping
together `separate` from `tidyr` and `mutate` from `dplyr`. `separate`
does exactly what its name implies: it separates a single character column
into multiple columns. `mutate` is used to add a new feature often (as
in this case) by acting on values of another feature.

## Listing 2.3 Feature normalization

The below code will normalize a feature using the "min-max" method. As
an example, the `Age` feature of the `titanic` dataframe is normalized
and a histogram of the new normalized feature is plotted with
[`ggplot2`](http://docs.ggplot2.org/current/index.html#). 

```{r normalize}
normalizeFeature <- function(data, fMin=-1.0, fMax=1.0){
  dMin = min(na.omit(data))
  dMax = max(na.omit(data))
  factor = (fMax - fMin) / (dMax - dMin)
  normalized = fMin + (data - dMin)*factor
  normalized
}

titanic$AgeNormalized <- normalizeFeature(titanic$Age)
ggplot(data=titanic, aes(AgeNormalized)) +
  geom_histogram()
```

## Figure 2.12 Mosaic plot for Titanic data: Gender vs. survival

The ["Visualizing Categorical Data" (`vcd`)](https://cran.r-project.org/web/packages/vcd/vcd.pdf) package 
provides an excellent set of functions for exploring categorical data,
including mosaic plots.

```{r figure2.12}
mosaic(
  ~ Sex + Survived,
  data = titanic, 
  main = "Mosaic plot for Titanic data: Gender vs. survival",
  shade = TRUE,
  split_vertical = TRUE,
  labeling_args = list(
    set_varnames = c(
      Survived = "Survived?")))
```

## Figure 2.13 Mosaic plot for Titanic data: Passenger class vs. survival

```{r figure2.13}
mosaic(
  ~ Pclass + Survived,
  data = titanic, 
  main = "Mosaic plot for Titanic data: Passenger Class vs. survival",
  shade = TRUE,
  split_vertical = TRUE,
  labeling_args = list(
    set_varnames = c(
      Pclass = "Passenger class",
      Survived = "Survived?")))
```

## Figure 2.14 Box plot for Titanic data: Passenger age vs. survival

The `boxplot` function is provided as part of the standard `graphics` package
in R. `ggplot2` provides a much nicer version.
```{r figure2.14a}
boxplot(Age ~ Survived, 
        data = titanic,
        xlab = "Survived?",
        ylab = "Age\n(years)",
        las = 1)
```

```{r figure2.14b}
ggplot(titanic, aes(Survived, Age)) + 
  geom_boxplot() +
  xlab("Survived?") +
  ylab("Age\n(years)")
```

## Figure 2.15 Box plots for Titanic data: Passenger fare versus survival

Plots can be combined in rows and columns using the `mfrow` graphical
parameter set via the `par` function. (Try `help(par)` to learn more.) 
```{r figure2.15}
par(mfrow=c(1,2))
par(mai=c(1,1,.1,.1), las = 1)
boxplot(Fare ~ Survived, 
        data = titanic,
        xlab = "Survived?",
        ylab = "Fare Amount")
boxplot(Fare**(1/2) ~ Survived, 
        data = titanic,
        xlab = "Survived?",
        ylab = "sqr\n(fare amount)")
```

## Figure 2.16 Density plot for MPG data, by region

The `sm.density.compare` function from the `sm` package is useful for
comparing a set of univariate density estimates.  The first argument is
a vector of data, and the second argument is a vector of group labels that
correspond to each value. The `colfill` variable and `legend` function
are used to place a legend on the plot generated by `sm.density.compare`.

```{r figure2.16}
par(mfrow=c(1,1))
auto <- read.csv("../data/auto-mpg.csv",
                 colClasses = c(origin = "factor"))

auto$origin <- revalue(auto$origin, 
                       c("1\t"="USA", "2\t"="Europe", "3\t"="Asia"))

sm.density.compare(auto$mpg, auto$origin,
                   xlab="Miles per gallon",
                   ylab="Density")
title(main="Density plot for MPG data, by region")

colfill<-c(2:(2+length(levels(auto$origin)))) 
legend(x="topleft", 
       inset=0.05, 
       text.width=5, 
       levels(auto$origin), 
       fill=colfill)
```

## Figure 2.17 Scatterplots for MPG data

It doesn't get much simpler than the `plot` function in R.
```{r figure2.17}
par(mfrow=c(1,2), mai=c(1,1,.1,.1), las = 1)
plot(auto$weight, auto$mpg,
     xlab = "Vehicle weight",
     ylab = "Miles per\ngallon")
plot(auto$modelyear, auto$mpg,
     xlab = "Model year",
     ylab = "Miles per\ngallon")

```

Although not in the book, everyone should become familiar with facets via 
[`ggplot2`](http://docs.ggplot2.org/current/index.html#).

```{r figure2.17_ggplot2}
p <- ggplot(auto, aes(mpg, weight)) + 
  geom_point()
p + facet_grid(. ~ origin)
```